An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2015-04-20 Response.txt

Sincerely yours,
Will Farr, Ilya Mandel, Daniel Stevens

----

Report 1:

>>> I am familiar with an earlier draft of this paper from internal LIGO review, and I'm pleased to see that it is finally being submitted for publication.

@@

...

>>> One minor quibble that I have is that the language in the introduction makes it seem that the underlying idea is different. Both methods "construct an approximation to each model’s posterior parameter distribution". The difference is that the kDE tree approach is far superior to  the original sparse histogram approach.

@@

We reformulated the sentence in question as follows to remove the ambiguity and make it clear that we are proposing an improved interpolation technique for constructing an approximate posterior, not a fundamentally different idea:

"...Here we introduce an alternative technique based on a kD-tree data  structure to construct an approximation to each model's posterior parameter distribution.  This improved interpolation method leads to dramatically faster convergence of RJMCMC sample chains."

>>> Other minor point is that the reference to parallel tempering does not point to the original paper on the topic (Swendsen and Wang 1986).

@@

We included the missing reference.

----

Report 2:

>>>This paper addresses an interesting and important issue in Bayesian model selection: Developing
efficient jump proposals in reversible-jump Markov-chain Monte Carlo (RJMCMC)
algorithms to transition between the models under consideration (high rejection rates for
inter-model proposals represent the limiting factor in many applications). The approach
proposed here|to exploit sampling of the posterior probability density (PPD) in the individual
model spaces, characterized as a kD tree structure|seems like a very promising idea
which could justify publication.

@@

We thank the referee for this assessment.

>>> However, there are a number of significant issues that must be addressed before publication is recommended.

We address these issues below, and in the text where appropriate.  We are very grateful to the referee for the extremely thorough reading of the text.

@@

The current paper does not repeat earlier publications, and includes key new material that has not been previously published.  The Farr et al., paper on "The Mass Distribution of Stellar-mass Black Holes" has been published in Astrophys. J. 741, 103, in 2011 (reference updated -- we apologize for the out-of-date reference which added to the confusion).  Farr+, 2011 dealt with measuring the mass distribution of stellar mass black holes in low-mass and high-mass X-ray binaries.  That analysis made use of an early implementation of the method described in this paper, and, consequently, a very brief summary of the method was included in Farr+, 2011.  However, the focus of Farr+, 2011, was on the astrophysical results rather than the method; consequently, the details of the algorithm and the thorough analysis presented here were not included in Farr+, 2011.  Moreover, the present paper includes a number of more recent (and unpublished) improvements to the method, such as the use of "rotated" boxes to improve the interpolation quality in higher-dimensional spaces (see Fig. 3 and associated text).  Finally, the use of the kD-interpolation scheme for adaptive sampling improvements in a single-model MCMC (Section 5 of the present paper) is entirely new.  Figure 4 illustrates the acceptance ratio of this single-model MCMC as applied to the very computationally challenging problem of parameter estimation on gravitational-wave parameter estimation (contrary to the referee's assertion that only toy models have been used for illustration).

>>> 2. The explanation of the inter-model proposal and acceptance criterion is Section 3c is
inadequate. While well-known material on Bayesian analysis (Section 3a) and MCMC
(Section 3b) is presented at length with equations, the new RJMCMC approach is
skimmed over in a single sentence (no equations) as “We perform MCMC in the supermodel
parameter space just like a regular MCMC; we propose jumps to different parameters
within a model (intra-model jumps) and jumps to a different model with
different parameters (inter-model jumps)" (page 6)...

We have added descriptive text at the bottom of page 5 and top of page 6 that expands on the RJMCMC approach.  The text now makes clear that the jump proposal ratio is a ratio of densities in the two model spaces, and discusses the special case where the spaces are of equal dimension and the proposal is a diffeomorphism between them (i.e., not a probabilistic choice of jump point, but rather a fixed mapping), where the ratio reduces to the Jacobian factor to which the referee refers.  We have not written the equation as given by the referee,

p_accept = (p(theta')*Q(theta' -> theta))/(p(theta)*Q(theta -> theta')) * |J|

because this form is not correct in our notation.  The factors Q *already incorporate the Jacobian* in our notation.  This point is now emphasised in the text.

We now give an explicit formula for the kD tree proposal density (Eq. 3.1 in the updated manuscript) with an expanded discussion of the proposal algorithm at the end of Section 3.  We have also modified the discussion surrounding Eq. 5.1 to account for the additional text now in Section 3.

@@

>>> 3. The examples given in the manuscript seem simplified to the point of triviality and/or
are poorly explained. The only example given of model selection (the potentially important
contribution of this paper) only involves sampling between 1D Gaussian and
Cauchy distributions (two unknown parameters each). While this does illustrate the
method (barely), it is of no practical relevance and is so much simpler than most realistic
model-selection problems that it is not clear how efficiency and performance
would scale, or how useful the method could be in practice. I assume the authors are
researching real-world model-selection problems of practical interest (like alternative
mass distribution models for black-hole X-ray binaries, mentioned on page 4). Including
a meaningful example such as this would greatly improve the paper and increase
its significance. The other two examples represent only single-model sampling and are
of much less interest, but even so they should be properly explained. In the examples
in Figures 3 and 4, what are the problems, what are the parameters, and what do the
results mean? This current presentation is not satisfying to the reader.

We have added text to the beginning of Section 4 pointing to the use of this algorithm in real-world applications, such as those of Farr et al (2011).  We feel that to describe the (lengthy) analysis from that work would needlessly complicate the presentation of the algorithm in this paper, but hope that this makes it clear that the algorithm is, indeed, applicable to complicated model selection problems.  We have publicly released the algorithm as a software library for any users interested in adopting it, and emphasize this point in the text.

We have added text to Figures 3 and 4 describing the gravitational-wave data analysis problem and the associated parameter space briefly.  Again, we feel that a complete description of this problem (available from Veitch et al (2015) and the other references in the text) is beyond the scope of the article, but we appreciate the request for more context in the explanation of this paper.  We hope the additional text provides this context.

>>> 4. Claims are made of “dramatically improved convergence" (page 3), that “convergence
is generally rapid" (page 4), and “the interpolation method described above can dramatically
improve the runtime of an RJMCMC" (page 10). While I believe the method
developed here could substantially improve efficiency in some cases, this is not actually
demonstrated in the paper. The trivial example in Figure 2 appears to increase
inter-model acceptance by a factor of 8 or so, but no results on RJMCMC sampling
convergence are given. The example considered in Figure 4 claims to have improved
convergence by a factor of two (not dramatic), but there is no discussion of what definition is applied for convergence|and as noted on page 6 (third paragraph), defining convergence for MCMC sampling is a tricky business. Even applying a reasonable
working definition of convergence, given the random nature of MCMC sampling, averages
over several MCMC runs are often required to meaningfully compare convergence
rates -- was that done here? While the claims of “dramatic" improvements in convergence may be true in some cases, they don't appear to be properly established in the paper.

We have removed the (perhaps over-wrought) adjective "dramatic" from our claims of convergence, and included text briefly describing the method of computing convergence times used and referring the reader for details to Veitch et al (2015).

@@

>>>  5a. Notation: While notation differs between disciplines, I am used to boldface (rather
than over-arrows) being used to represent vectors in published work -- the authors
could consider this. Further, why is the set of parameters written as a vector,
but the set of data not written as a vector (d)?

We write parameters as a vector because we occasionally consider individual parameters or subsets of the parameter set; no such confusion is likely for the data.

>>> 5b. Page 5, Frst line after Section 2b heading: Should be j = 1…; not i = 1…;

Fixed.

>>> 5c. Page 5, third line after Section 2b heading: Should be “set of parameters" not
“set of parameter".

Fixed.

>>> 5d. Figure 1 would be clearer if parameter axes were added indicating the origin, with
axes scales in terms of standard deviations in x and y.

After attempting to follow the referee’s suggestion, we came to the conclusion that it distracted from the clarity of the illustration of kD interpolation.

>>> 5e. The x axis of Figure 2 is labeled Nboxing while it should be Nbox.

Nboxing is the minimum number of samples per box used to evaluate the local density, while Nbox is the actual number of samples (can be different since the total number of samples is not an exact power of 2); see Section 3.  Nbox is thus specific to a particular point, while Nboxing is a tuning parameter of the algorithm and is the parameter being varied in Fig. 2.

>>> 5f. Figure 3 caption: kd should be kD.

Fixed.

>>> 5g. Equation (5.1): P should be Q (this is the proposal distribution). Also, V might
be better written as Vbox to indicate that it is the volume of the current box, like
Nbox indicates the number of samples in the current box (V is not a constant but
varies with boxes).

Fixed here and in subsequent paragraph.  We kept V, since we define it explicitly just after this equation (and it has a slightly different meaning here then in (3.1).

>>> 5h. Page 11: Inconsistent use of math or Roman font for some subscripts; e.g., Nbox
and Nbox, and Ncrit. When the subscript is a label rather than an index, Roman
font is preferred.

Fixed.

>>> 5i. Page 11, after Eq. (5.1): Please give an equation for V , the volume of the intersection
of the two (potentially high-dimensional) boxes.

We have omitted this as the expression is somewhat tedious without adding any insight.

>>> 5j. Page 12, first paragraph discusses violations of detailed balance arising from using
the sampling history to define proposals. Such methods are well known, including
the requirements which must be met to avoid biased sampling (so-called diminishing
adaptations). References (such as Brooks et al., 2011 Handbook of Markov
chain Monte Carlo, New York: Springer) should be included here to be clear this
is not a new idea.

We did not wish to imply that this is a new idea; suggested reference added.

>>> 5k. Figure 4's caption states “The acceptance probability rate asymptotes to the
steady-state solution once sufficient samples are accumulated for the kD tree
to allow the sample density or be accurately interpolated." This asymptotic
convergence appears to occur by about step 2.0 * 10^6; hence, all samples collected
prior to this point should be discarded as burn-in or the sampling is potentially
biased. This should be mentioned explicitly in the paper.

It is now clarified:
“The acceptance rate asymptotes to the steady-state solution once sufficient samples have been accumulated in the kD tree to allow the sample density to be accurately interpolated; samples collected
prior to this point should be discarded as burn-in.”

>>> 5l. Figure 4: Why denote the x axis divisions from 0 to 25 000 and then include
\( X100)" in the axis label? Why not denote the divisions from 0 to 2.5 and
include \(X 10^6)" in the label|this is the common use of exponential notation.

The referee’s suggestion would, indeed, be more standard, but the tools we used to process the data made it more straightforward to generate the plot as given, so we hope the slightly odd axis label is not too jarring.

### 2015-04-20 paper.tex
Updates to include response to items from 5b...l from report by referee 2.

\newcommand{\vtheta}{\vec{\theta}}
\newcommand{\Nbox}{N_\mathrm{box}}
\newcommand{\Vbox}{V_\mathrm{box}}

@@

A Markov chain Monte Carlo \cite{Gilks:1996} produces a set of samples
$\{ \vtheta^{(j)} \, | \, i = 1, \ldots \}$ from the model parameter
space that are sampled according to the posterior, meaning that, in
the limit that the chain length tends to infinity, the relative
frequency with which a given set of parameter appears in the chain is
proportional to the desired posterior, $p(\vtheta|d,M)$.  Therefore,
the output of an MCMC can be directly interpreted as the posterior PDF
over the full parameter space, while PDFs for individual parameters
can be obtained by marginalizing over the uninteresting parameters.

@@

The kD-tree based interpolated jump proposal described in Section \ref{sec:kDTree} selects one sample from the tree at random and proposes a uniform point from the box containing that sample.  This ignores the density of other samples in the neighborhood of this box, which contains additional information about the true density within the box, which is likely non-uniform.  A better proposal, therefore, would account for larger-scale sample covariances in parameter space.  To address significant covariance between samples, which would correspond to strong density gradients and a very non-uniform density profile within a rectangular box, the modified jump proposal does not descend to the leaf-nodes of the tree, but instead stops descending whenever the number of samples in the current box, $N_{box}$, falls below some threshold, $N_{crit}$. We then find the principal axes of the samples
contained in this box by calculating the eigenvectors of the covariance matrix of these samples. We use these eigenvectors to draw a new box. This box -- henceforth called a covariance cell -- contains all $N_{box}$ samples from the kD-tree box, is centered on the mean of these samples, and has edges that are aligned with the principal axes. Furthermore, this box is drawn around the samples as tightly as possible; in other words, for each edge, there exists a point in the covariance cell that lies along that edge. Figure \ref{fig:PCC} illustrates these two boxes.

\begin{figure}
  \includegraphics[width=0.8\columnwidth]{PCC}
  \caption{\label{fig:PCC} A two-dimensional illustration of the two
    boxes involved in the modified kD proposal.  The larger box
    aligned with the axes is the normal kD-tree box containing the
    given samples, while the tighter box is the ``covariance cell''
    aligned with the eigenvectors of the sample distribution.  The
    modified kd-interpolated jump proposal draws a point from the
    intersection of the two boxes. Tightly-correlated posteriors in
    parameter space such as this are typical of the gravitational-wave
    parameter estimation problem described in \cite{Veitch:2014} and
    in the text.  Without the modification to account for the
    correlated samples, the kD neighbourhood of these points would
    produce a very inefficient jump proposal, since most of the
    bounding-box would contain empty (i.e.\ low-posterior) parameter
    space.}
\end{figure}

The jump proposal chooses a stored sample at random and finds the largest kD-tree box containing it and fewer than $N_{crit}$ total samples. It then draws the covariance cell and proposes a new point uniformly from the intersection of the covariance cell with the original kD-tree box.
The jump probability for this jump proposal is still (c.f.\ Eq.\
\eqref{eq:neighbourhood-jp}) given by
%
\bel{modforward} P(\vtheta_i \rightarrow \vtheta_{i+1}) =
\frac{\Nbox}{N V} , \ee
%
where $N$ is the total number of samples in the kD tree, and $V$ is the
volume of the intersection between the kD-tree box and the correlation cell containing $\vtheta_{i+1}$. When used as the only jump proposal for an
RJMCMC, it is important that this proposal be capable of
proposing points in all allowed regions of parameter space; on the
other hand, when the kD proposal is used in a suite of other jump proposals, this is no longer a requirement.

@@

This technique trivially violates detailed balance since the jump proposal distribution depends on the past history. One way to address this problem requires continuously accumulating samples into the kD tree as the MCMC explores more
of the parameter space. As the number of samples drawn from the posterior PDF increases, the neighborhoods around each sample decrease with increasing sample density and the calculated covariances between samples in the kD tree better approximate the true covariances between parameters in the equilibrium posterior PDF.  Hence, for large $N$, the change in $P(\vtheta^{(j)} \rightarrow \vtheta^{(j+1)})$ approaches zero and the jump proposal does a progressively better job of sampling the equilibrium posterior while becoming asymptotically Markovian.  Conservatively, the entire set of samples obtained while the jump proposal is being dynamically updated could be discarded as burn-in, and future analysis could use the static jump proposal distribution interpolated from samples accumulated during the extended burn-in phase. 
%Because this introduces a dependence on the
%previous states of the MCMC, the jump proposal is no longer
%Markovian.  However, it remains \emph{asymptotically} Markovian \ilya{[And being asymptotically Markovian (whatever that actually means -- at least a reference?) -- is provably sufficient to guarantee that we sample from the equilibrium distribution even though we nominally violate detailed balance? (again, a reference?)]} , which
%ensures that the equilibrium distribution of the MCMC is correct.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{linearlog3}
  \end{center}
  \caption{\label{fig:accratio} The cumulative acceptance
    ratio for the modified kD-tree jump proposal used in the
    \texttt{LALInferenceMCMC} code as a function of the number of
    steps (in hundreds). The simulation in question was a
    nine-dimensional analysis of a simulated gravitational-wave signal
    injected into synthetic data similar to that taken by the LIGO
    and Virgo detectors \cite{AdvLIGO,AdvVirgo}.  The parameter space
    includes the masses of the compact objects generating the
    gravitational-wave, their location on the sky, distance, orbital
    orientation, the time of signal arrival, and the orbital phase.
    The posterior in this problem has a number of well-separated modes
    in the parameter space which are difficult to jump between using
    traditional jump proposals; in spite of the small acceptance ratio
    of the kD proposal, when applied to one in twenty jumps proposed
    in this simulation, it improved the convergence time of the sampler by a factor of two
    compared to using only the standard suite of proposals. 
    The acceptance rate asymptotes to the steady-state solution once sufficient samples have been accumulated in the kD tree to allow the sample density to be accurately interpolated.
%\dan{[Taking a leap here, someone please confirm/deny/add qualifiers] The initial decrease in acceptances arises from insufficient population of the kD tree with too few samples, while the leveling-off later in the MCMC indicates convergence.} \ilya{[I see structure in this figure (a high acceptance probability, followed by a dip, followed by an asymptote).  Is this a fluke, or do you always see this -- and, if so, why?  BTW, what's the autocorrelation length (just to give me a sense of scale, so I am not reading too much into meaningless fluctuations within an AC length)?]} \dan{[Don't recall the autocorrelation length. Will?]}
  }
\end{figure}

>>>>

\newcommand{\vtheta}{\vec{\theta}}
\newcommand{\Nbox}{N_\mathrm{box}}
\newcommand{\Ncrit}{N_\mathrm{crit}}
\newcommand{\Vbox}{V_\mathrm{box}}

@@

A Markov chain Monte Carlo \cite{Gilks:1996} produces a set of samples
$\{ \vtheta^{(j)} \, | \, j = 1, \ldots \}$ from the model parameter
space that are sampled according to the posterior, meaning that, in
the limit that the chain length tends to infinity, the relative
frequency with which a given set of parameters appears in the chain is
proportional to the desired posterior, $p(\vtheta|d,M)$.  Therefore,
the output of an MCMC can be directly interpreted as the posterior PDF
over the full parameter space, while PDFs for individual parameters
can be obtained by marginalizing over the uninteresting parameters.

@@

The kD-tree based interpolated jump proposal described in Section \ref{sec:kDTree} selects one sample from the tree at random and proposes a uniform point from the box containing that sample.  This ignores the density of other samples in the neighborhood of this box, which contains additional information about the true density within the box, which is likely non-uniform.  A better proposal, therefore, would account for larger-scale sample covariances in parameter space.  To address significant covariance between samples, which would correspond to strong density gradients and a very non-uniform density profile within a rectangular box, the modified jump proposal does not descend to the leaf-nodes of the tree, but instead stops descending whenever the number of samples in the current box, $\Nbox$, falls below some threshold, $\Ncrit$. We then find the principal axes of the samples
contained in this box by calculating the eigenvectors of the covariance matrix of these samples. We use these eigenvectors to draw a new box. This box -- henceforth called a covariance cell -- contains all $\Nbox$ samples from the kD-tree box, is centered on the mean of these samples, and has edges that are aligned with the principal axes. Furthermore, this box is drawn around the samples as tightly as possible; in other words, for each edge, there exists a point in the covariance cell that lies along that edge. Figure \ref{fig:PCC} illustrates these two boxes.

\begin{figure}
  \includegraphics[width=0.8\columnwidth]{PCC}
  \caption{\label{fig:PCC} A two-dimensional illustration of the two
    boxes involved in the modified kD proposal.  The larger box
    aligned with the axes is the normal kD-tree box containing the
    given samples, while the tighter box is the ``covariance cell''
    aligned with the eigenvectors of the sample distribution.  The
    modified kD-interpolated jump proposal draws a point from the
    intersection of the two boxes. Tightly-correlated posteriors in
    parameter space such as this are typical of the gravitational-wave
    parameter estimation problem described in \protect\cite{Veitch:2014} and
    in the text.  Without the modification to account for the
    correlated samples, the kD neighbourhood of these points would
    produce a very inefficient jump proposal, since most of the
    bounding-box would contain empty (i.e., low-posterior) parameter
    space.}
\end{figure}

The jump proposal chooses a stored sample at random and finds the largest kD-tree box containing it and fewer than $\Ncrit$ total samples. It then draws the covariance cell and proposes a new point uniformly from the intersection of the covariance cell with the original kD-tree box.
The jump probability for this jump proposal is still (c.f.\ Eq.\
\eqref{eq:neighbourhood-jp}) given by
%
\bel{modforward} Q(\vtheta_i \rightarrow \vtheta_{i+1}) =
\frac{\Nbox}{N V} , \ee
%
where $N$ is the total number of samples in the kD tree, and $V$ is the
volume of the intersection between the kD-tree box and the correlation cell containing $\vtheta_{i+1}$. When used as the only jump proposal for an
RJMCMC, it is important that this proposal be capable of
proposing points in all allowed regions of parameter space; on the
other hand, when the kD proposal is used in a suite of other jump proposals, this is no longer a requirement.

@@

This technique trivially violates detailed balance since the jump proposal distribution depends on the past history. One way to address this problem, through diminishing adapations \cite{Brooks:2011}, requires continuously accumulating samples into the kD tree as the MCMC explores more
of the parameter space. As the number of samples drawn from the posterior PDF increases, the neighborhoods around each sample decrease with increasing sample density and the calculated covariances between samples in the kD tree better approximate the true covariances between parameters in the equilibrium posterior PDF.  Hence, for large $N$, the change in $Q(\vtheta^{(j)} \rightarrow \vtheta^{(j+1)})$ approaches zero and the jump proposal does a progressively better job of sampling the equilibrium posterior while becoming asymptotically Markovian.  Conservatively, the entire set of samples obtained while the jump proposal is being dynamically updated could be discarded as burn-in, and future analysis could use the static jump proposal distribution interpolated from samples accumulated during the extended burn-in phase. 
%Because this introduces a dependence on the
%previous states of the MCMC, the jump proposal is no longer
%Markovian.  However, it remains \emph{asymptotically} Markovian \ilya{[And being asymptotically Markovian (whatever that actually means -- at least a reference?) -- is provably sufficient to guarantee that we sample from the equilibrium distribution even though we nominally violate detailed balance? (again, a reference?)]} , which
%ensures that the equilibrium distribution of the MCMC is correct.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{linearlog3}
  \end{center}
  \caption{\label{fig:accratio} The cumulative acceptance
    ratio for the modified kD-tree jump proposal used in the
    \texttt{LALInferenceMCMC} code as a function of the number of
    steps (in hundreds). The simulation in question was a
    nine-dimensional analysis of a simulated gravitational-wave signal
    injected into synthetic data similar to that taken by the LIGO
    and Virgo detectors \protect\cite{AdvLIGO,AdvVirgo}.  The parameter space
    includes the masses of the compact objects generating the
    gravitational-wave, their location on the sky, distance, orbital
    orientation, the time of signal arrival, and the orbital phase.
    The posterior in this problem has a number of well-separated modes
    in the parameter space which are difficult to jump between using
    traditional jump proposals; in spite of the small acceptance ratio
    of the kD proposal, when applied to one in twenty jumps proposed
    in this simulation, it improved the convergence time of the sampler by a factor of two
    compared to using only the standard suite of proposals. 
    The acceptance rate asymptotes to the steady-state solution once sufficient samples have been accumulated in the kD tree to allow the sample density to be accurately interpolated; samples collected
prior to this point should be discarded as burn-in.
  }
\end{figure}
